---
title: "Data_Analysis"
author: "Idris Iritas"
date: "2024-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis

In this notebook we use some data visualisation tools to explore a dataset of Albert Heijn products

```{r, echo=FALSE}
# Imports
library(ggplot2)

```

```{r}
# Get project root paths
r_project_root <- getwd()
main_project_root <- dirname(r_project_root)

```

Load in the dataset from csv. The dataset used here is provided with the repository.

```{r}
data_path <- file.path(main_project_root, "complete_datasets", "2024-10-23.csv")

df <- read.csv(data_path)

dim(df)
```

This dataset has 30498 rows, or products, and 106 columns, or data features.

What these features are is showcased below

```{r}
colnames(df)
```

A separate column was created for every nutrient encountered during scraping. This makes the dataframe unnecessarily large, as we do not use most of this data. We'll filter out all nutrients except for Energy values, carbohydrates, sugars, saturated fats, unsaturated fats, total fats, protein, fiber and salt.

```{r}
df <- df[, -c(22:99)]

dim(df)
```

Now we have 28 columns of data. Next we want to be aware of the number of NA values for each column. We have a lot of missing data, such as missing nutrition for non-food items, missing categories and missing ingredients. Let's count the missing values per feature:

```{r}
na_counts <- sapply(df, function(x) sum(is.na(x)))

na_counts
```

```{r}

# Convert the named vector to a dataframe
na_df <- data.frame(Names = names(na_counts), Values = as.numeric(na_counts))

# Create the bar plot, sorted by Values in descending order
ggplot(na_df, aes(x = reorder(Names, -Values), y = Values)) +
  geom_bar(stat = "identity", fill = "darkred") +
  coord_flip() + # Flip the plot for better readability
  labs(x = "", y = "NA Values Count")
```

We observe that there is not much variation in the NA count for the nutrients: Energy, Carbohydrates, Sugars, Fats, Saturated fats, Protein and Salt. We can create a data-frame for values that have these nutrients labeled. But we have to do some data cleaning. The data is suffixed with units like "g" and "mg" and sometimes prefixed with "\<". Let's check which units exist for the nutritional columns

```{r}
# Function to check if only 'g' is present
check_units <- function(column) {
  
  # Extract unique units using regular expression
  units <- unique(gsub(".*?(g|kg|mg|l).*", "\\1", column))
  return(unique(units))
  
}

# Apply the function to each nutritional column
result <- sapply(df[, c(14:21)], check_units)

# Print results
print(result)
```

We observe that g is the only unit used for all columns except for salt.

Let's parse all the columns with only g, then handle the salt column afterwards:

```{r}
# Store these values for a sanity check
old_values <- df$Koolhydraten[c(4328, 4619, 12249, 12697, 23202, 21876)]


# Loop through columns 14 to 20
for (col in 14:20) {
  # Remove " g" and handle special cases like "< something g"
  cleaned_values <- gsub(" g", "", df[[col]])
  
  # Replace any values that start with "<" with NA
  cleaned_values <- ifelse(grepl("^<", cleaned_values), 0, cleaned_values)
  
  # Convert to numeric
  df[[col]] <- as.numeric(cleaned_values)
}

# Store these values for a sanity check
new_values <- df$Koolhydraten[c(4328, 4619, 12249, 12697, 23202, 21876)]
```

To check the result of our processing, we took a few indices as test-cases:

```{r}
aligned_output <- data.frame(Raw = old_values, Processed = new_values)

aligned_output
```

It is important to note that with our rule-set of converting \< x to 0, we also convert \<19 g to 0.

Now to handle the salt column, we perform the same parsing operation as before, except now we also divide by 1000 if we encounter mg to have all data as the same unit: grams.

```{r}
old_salt_values <- df$Zout[c(23254, 263, 9090, 7928, 19921, 2, 250)]

# Remove " g" and handle special cases like "< something g" and milligrams
cleaned_values <- gsub(" mg", "", df$Zout)   # Remove " mg" for easier processing
cleaned_values <- gsub(" g", "", cleaned_values)  # Remove " g" for easier processing

# Replace any values that start with "<" with 0
cleaned_values <- ifelse(grepl("^<", cleaned_values), 0, cleaned_values)

# Convert to numeric and handle mg
df$Zout <- ifelse(grepl("mg$", df$Zout), as.numeric(cleaned_values) / 1000, as.numeric(cleaned_values))

new_salt_values <- df$Zout[c(23254, 263, 9090, 7928, 19921, 2, 250)]
```

We can perform another sanity check below, and we see that the values were all converted correctly.

```{r}
aligned_salt_output <- data.frame(Raw = old_salt_values, Processed = new_salt_values)
aligned_salt_output$Processed <- sprintf("%.4f", aligned_salt_output$Processed)

aligned_salt_output
```

Now that our numerical nutrition data is processed, let's create a new data-frame that excludes all NA values for our aforementioned nutrients.

```{r}
# Create the subset of df with the specified columns
nutrition_df <- df[, c("ProductId", "ProductName", "Energie..kcal.", "Vet", 
                        "waarvan.verzadigd", "Koolhydraten", 
                        "waarvan.suikers", "Eiwitten", "Zout")]

nutrition_df <- na.omit(nutrition_df)

head(nutrition_df)

```

Before we continue, I noticed a discrepancy between kcal and kJ energy measurements in the nutrition data:

```{r}
error_indices <- !is.na(df$Energie..kcal.) & !is.na(df$Energie..kJ.) & (df$Energie..kcal. > df$Energie..kJ.)

error_products <- df[error_indices, c("ProductName", "Energie..kcal.", "Energie..kJ.")]

print(error_products)

```
